Non-word repetition in Yélî Dnye

### 2021-01-06 manuscript.Rmd
done going through results, still in progress

The final list includes: three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).

```{r tab2-stims, results = "asis"}
#add orthographic representations targets & other stims chars
stims.tbl <- read.xlsx("stimuli.xlsx") %>%
  arrange(factor(type, levels = c("practice", "1", "2", "3", "4"))) %>%
  select(type, target_ortho, phono) %>%
  rename("Type" = type, "Orthographic" = target_ortho, "Phonological" = phono)

@@

Finally, when describing children's patterns of errors, we take all repetitions of a given target into account. We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, insertions, and substitutions.

@@

@@

#check that indeed:
#all numbers ... sum to 100 across sub, del, mat at the TARGET level 
#NOTE! if you round above, then these don't sum to 100, so don't round very much...
#rowSums(sub_pc, na.rm=T) + del_pc + mat_pc
```

Next, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data. Deletions were very rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets). We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1])*100,2)` of the `r dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1]` phones found collapsing across all children and target words, so that substitutions constituted the frank majority of incorrect phones (~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] )*100,2)` of unmatched phones).

<!-- reminder: this table has target as rows, realization as columns  -->
Is it the case that cross-linguistically less common phones are more frequently substituted by more common ones than vice versa? We found that errors involving nasal vowel targets tended to result in a change in nasality (`r sum(sub_tab[nasalVs,oralVs])` nasal targets produced as oral, ~`r round(sum(sub_tab[nasalVs,oralVs])/sum(sub_tab[nasalVs,])*100)`% of substitutions for nasal vowel target phones) more frequently than in a change in quality (`r sum(sub_tab[nasalVs,nasalVs])` nasal targets produced as a different nasal vowel, ~`r round(sum(sub_tab[nasalVs,nasalVs])/sum(sub_tab[nasalVs,])*100)`% of substitutions for nasal targets). The opposite was true for oral vowels: there were `r sum(sub_tab[oralVs,nasalVs])` oral targets produced as nasal, constituting ~`r round(sum(sub_tab[oralVs,nasalVs])/sum(sub_tab[oralVs,])*100)`% of substitutions for oral target phones; compared to `r sum(sub_tab[oralVs,oralVs])` oral targets produced as a different oral vowel, ~`r round(sum(sub_tab[oralVs,oralVs])/sum(sub_tab[oralVs,])*100)`% of substitutions for oral targets).

**STOPPED HERE**

Among consonants, complex consonants were occasionally substituted by simpler ones (specifically `r print.phone.list(complexCs.ipa)` as `r print.phone.list(simplexCs.ipa)`), `r sum(sub_tab[complexCs,simplexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all complex consonant targets. Interestingly, simple consonants were sometimes also substituted by complex ones; this happened `r sum(sub_tab[simplexCs,complexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all simple consonant targets. Finally, simple consonants were mispronounced as other simple consonants (`r sum(sub_tab[simplexCs,simplexCs])` instances, ~`r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% of all simple consonant targets) and complex consonants as other complex consonants (`r sum(sub_tab[complexCs,complexCs])` instances, ~`r mean(sub_pc[complexCs,complexCs],na.rm=T)`% of all complex consonant targets). Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.

Finally, we looked into the frequency with which mispronunciations resulted in real words. In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%. This type of analysis is seldom reported. We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases. The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.

@@

aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$type),mean)->by_length_no_child
colnames(by_length_no_child)<-c("Length","Word")
by_length_no_child=by_length_no_child[-1,]
length_r=cor.test(by_length_no_child$Length,by_length_no_child$Word)

@@

Next, we inspected whether NWR scores varied as a function of word length (Table \@ref(tab:tablength)). Participants scored much lower on monosyllables than on non-words of other lengths. This is likely due to the fact that the majority of monosyllables were designed to include sounds that are rare in the world's languages, which may indicate that they are hard to produce or perceive. Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole-item scoring. While whole-item scoring is the most commonly reported NWR outcome, it is also the least forgiving. Averaging across participants and items, the Pearson correlation between length (2-4 syllables) and whole-item NWR scores was r(`r length_r$parameter`) = `r round(length_r$estimate,3)`. In a generalized binomial mixed model, we included `r summary(len_lm)$devcomp$dims["N"]` observations, from `r summary(len_lm)$ngrps["file"]` children producing, in any given trial, one of `r summary(len_lm)$ngrps["target"]` potential target words. The analysis revealed a positive effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std. Error"]`, p `r pval.display(mycoeff["age.rounded_s","Pr(>|z|)"])`) and a negative but non-significant estimate for target length in number of syllables (ß = `r mycoeff["type","Estimate"]`, SE ß = `r mycoeff["type","Std. Error"]`, p `r pval.display(mycoeff["type","Pr(>|z|)"])`).

```{r tablength, results = "asis"}
length.tab=NULL
for(myvar in c("Word","NLD")){
  length.tab=cbind(length.tab,
              paste0(round(aggregate(by_length[,myvar],by=list(by_length$Length),mean)$x),
              " (",round(aggregate(by_length[,myvar],by=list(by_length$Length),sd)$x),")"))
}
rownames(length.tab)<-levels(by_length$Length)
colnames(length.tab)<-c("Word","NLD")
apa_table(length.tab,caption="NWR measured in whole-word scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
```

@@

@@

plot(overall$Word~overall$Age,pch=20,cex=2,lty=3,ylab="NWR scores (% words)",xlab="Age (years)",col=ifelse(overall$Sex=="B","darkmagenta","chocolate1"))

cor.test(overall$Word,overall$Age,method="s")->corage
t.test(overall$Word~overall$Sex)->tsex
cor.test(overall$Word,overall$BrthO,method="s")->corbo
cor.test(overall$Word,overall$matEd,method="s")->corme

@@

Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals. As shown in Figure \@ref(fig:Fig2-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`). In contrast, there was no clear association between NWR scores and sex (t (`r tsex$statistic`) = `r tsex$parameter`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r corbo$statistic`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (data missing for `r sum(is.na(demo$matEd[demo$included=="yes"]))` children,  rho ( `r corme$statistic`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).

@@

With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs. oral vowels and simple vs. complex consonants, revealed no remarkable bias in error types. That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically. Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus. Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work [e.g., @castro1998illiterate], but it is unclear what caused this pattern in the current study: Castro and colleagues' [-@castro1998illiterate] study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors. Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.

@@

>>>>

The final list includes: three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \@ref(tab:tabstims)).

```{r tabstims, results = "asis"}
#add orthographic representations targets & other stims chars
stims.tbl <- read.xlsx("stimuli.xlsx") %>%
  arrange(factor(type, levels = c("practice", "1", "2", "3", "4"))) %>%
  select(type, target_ortho, phono) %>%
  rename("Type" = type, "Orthographic" = target_ortho, "Phonological" = phono)

@@

@@

We also looked into the frequency with which mispronunciations resulted in real words. In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%. This type of analysis is seldom reported. We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases. The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.

@@

#check that indeed:
#all numbers ... sum to 100 across sub, del, mat at the TARGET level 
#NOTE! if you round above, then these don't sum to 100, so don't round very much...
#rowSums(sub_pc, na.rm=T) + del_pc + mat_pc

Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa? To answer this question, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data. Deletions were very rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets). We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1])*100,2)` of the `r dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1]` phones found collapsing across all children and target words, so that substitutions constituted the frank majority of incorrect phones (~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] )*100,2)` of unmatched phones).

```{r tab-v,results="asis"}

#vowel changes
mytab=rbind(
  #OK --                      nasal change --                 qual change
  cbind(sum(match_tab[nasalVs]),sum(sub_tab[nasalVs,oralVs]),sum(sub_tab[nasalVs,nasalVs])), #nasal target
  cbind(sum(match_tab[oralVs]),sum(sub_tab[oralVs,nasalVs]),sum(sub_tab[oralVs,oralVs])) #oral target
)
mytab=data.frame(mytab)
mytab$pc_corr=round(mytab[,1]/rowSums(mytab)*100,1)
mytab$pc_nas_err=round(mytab[,2]/rowSums(mytab)*100,1)
mytab$pc_qual_err=round(mytab[,3]/rowSums(mytab)*100,1)
rownames(mytab)<-c("Nasal Target", "Oral Target")
colnames(mytab)<-c("Correct","Nasal Err.","Qual. Err","% Corr.","% Nasal Err.","% Qual Err.")

kable(mytab,
  "latex", booktabs = T,
  caption="Number (and percent) of vowel targets that were correctly repeated or substituted, as a function of vowel type, and whether the error resulted in a nasality change (Nasal Err.) or only a quality change (Qual. Err.) Deletion errors are not shown.")

We approached our question regarding potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not separately for nasal and oral vowels, distinguishing errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (but were therefore a quality error), shown in Table \@ref(tab:tab-v). We found that errors involving nasal vowel targets were more common than those involving oral vowels (`r round(100-mytab["Nasal Target","% Corr."],1)` versus `r round(100-mytab["Oral Target","% Corr."],1)`). Additionally, errors in which a nasal vowel lost its nasal character were 10 times more common than those in which an oral vowel was produced as a nasal one.

```{r tab-c,results="asis"}

#consnant changes
mytab=rbind(
  #OK --                      complexity change --                 other change
  cbind(sum(match_tab[complexCs]),sum(sub_tab[complexCs,simplexCs]),sum(sub_tab[complexCs,complexCs])), #nasal target
  cbind(sum(match_tab[simplexCs]),sum(sub_tab[simplexCs,complexCs]),sum(sub_tab[simplexCs,simplexCs])) #oral target
)
mytab=data.frame(mytab)
mytab$pc_corr=round(mytab[,1]/rowSums(mytab)*100,1)
mytab$pc_comp_err=round(mytab[,2]/rowSums(mytab)*100)
mytab$pc_oth_err=round(mytab[,3]/rowSums(mytab)*100,1)
rownames(mytab)<-c("Complex Target", "Simple Target")
colnames(mytab)<-c("Correct","Compl. Err.","Other Err","% Corr.","% Compl. Err.","% Other Err.")

kable(mytab,
  "latex", booktabs = T,
  caption="Number (and percent) of consonant targets that were correctly repeated or substituted, as a function of the complexity of the consonant, and whether the error resulted in a change of complexity (Compl Err.) or not (Other Err.) Deletion errors are not shown.")

For consonants, we inspected complex (`r print.phone.list(complexCs.ipa)`) versus simpler ones (`r print.phone.list(simplexCs.ipa)`), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.^[Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.] Results, shown in Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving oral vowels (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`). Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.

@@

@@

Next, we inspected whether NWR scores varied as a function of word length (Table \@ref(tab:tablength)). Participants scored much lower on monosyllables than on non-words of other lengths. This is likely due to the fact that the majority of monosyllables were designed to include sounds that are rare in the world's languages, which may indicate that they are hard to produce or perceive. Setting monosyllables aside, we observed the typical pattern of lower scores for longer items only for the whole-item scoring, and even there differences were rather small. In a generalized binomial mixed model excluding monosyllables, we included `r summary(len_lm)$devcomp$dims["N"]` observations, from `r summary(len_lm)$ngrps["file"]` children producing, in any given trial, one of `r summary(len_lm)$ngrps["target"]` potential target words. The analysis revealed a positive effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std. Error"]`, p `r pval.display(mycoeff["age.rounded_s","Pr(>|z|)"])`) and a negative but non-significant estimate for target length in number of syllables (ß = `r mycoeff["type","Estimate"]`, SE ß = `r mycoeff["type","Std. Error"]`, p `r pval.display(mycoeff["type","Pr(>|z|)"])`).

```{r tablength, results = "asis"}
length.tab=NULL
for(myvar in c("Word","NLD")){
  length.tab=cbind(length.tab,
              paste0(round(aggregate(by_length[,myvar],by=list(by_length$Length),mean)$x),
              " (",round(aggregate(by_length[,myvar],by=list(by_length$Length),sd)$x),")"))
}
rownames(length.tab)<-paste(c(1:4),"syll")
colnames(length.tab)<-c("Word","NLD")
kable(length.tab,"latex", booktabs = T,caption="NWR means (and standard deviations) measured in whole-word scores and normalized Levenshtein Distance (NLD), separately for the four stimuli lengths.")
```

@@

```{r indif}
cor.test(overall$Word,overall$Age,method="s")->corage
t.test(overall$Word~overall$Sex)->tsex
cor.test(overall$Word,overall$BrthO,method="s")->corbo
cor.test(overall$Word,overall$matEd,method="s")->corme
```

@@

plot(overall$Word~overall$Age,pch=4,cex=2,lty=3,ylab="NWR scores (% words)",xlab="Age (years)",col=ifelse(overall$Sex=="B","darkmagenta","chocolate1"))

@@

Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals. As shown in Figure \@ref(fig:Fig2-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`). In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (data missing for `r sum(is.na(demo$matEd[demo$included=="yes"]))` children,  rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).

@@

With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs. oral vowels and simple vs. complex consonants, revealed no remarkable bias in error types. That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.

@@

### Other observations

Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus. Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work [e.g., @castro1998illiterate], but it is unclear what caused this pattern in the current study: Castro and colleagues' [-@castro1998illiterate] study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors. Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.
